Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-04-01 WSSSPE1_report/WSSSPE1_summary_report.tex
some text changes

One group of papers presented frameworks that were primarily about characteristics of software artifacts, connecting with the long discourse on software quality. This approach is realized in adjectives that can be applied to pieces of software but might also extend to describe software projects.  Thus Calero and colleagues~\cite{Calero_WSSSPE} propose adding elements to the ISO standards for measuring software quality: proposing an additional dimension of quality they call perdurability or ability to ``last over time'' made up of three characteristics: reliability, maintainability and adaptability. This shows some overlap with the framework employed by Venters et al.~\cite{Venters_WSSSPE} who employ the adjectives of ``extensibility, interoperability, maintainability, portability, reusability and scalability,'' anticipating the sorts of work that people would need to do with a software artifact in the future, ``as stakeholders requirements, technology and environments evolve and change.'' Venters et al. argue that these choices are really made very early because they relate to the architecture of the software and involve trade offs that ought to be analyzed alongside other attributes of the software artifact. Lenhardt et al~\cite{Lenhardt_WSSSPE} compare the software lifecycle to the data lifecycle to argue for the inclusion of metadata throughout a piece of software's life (discussing, for example, how it has been built and tested and what ``data in'' and ``data out'' has been considered). In addition, their analogy suggests that the software lifecycle might add a phase of ``preservation'' and draw on understandings of what that involves from studies of data.  In sum, then, these frameworks focused on what needs to be accomplished to have more sustainable software.

A second theme in these papers was the continued availability of resources to accomplish the goals of sustainability. The elements of these frameworks focused far more on the organization of a software project than they did on characteristics of the artifact itself (although it is certainly true that the adjectives discussed above could be applied to a software project). For example Pierce et al~\cite{Pierce_WSSSPE} focus on the way that the project is run, particularly in terms of how those involved communicate and jointly set priorities, a process they call governance. In particular they argue that because sustainability relates to having ongoing resources, governance must be open to receive diverse input (by occurring online, asynchronously, and publicly) and thus have the potential to ``transform passive users into active contributors''. They argue that the Apache Software Foundation's incubation process teaches this and could be learned from by projects throughout scientific software.  Katz and Proctor~\cite{Katz_WSSSPE} also discuss governance, describing two modes: ``top-down'' and ``bottom-up'' governance.  They place governance alongside questions of the artifact (``technical''), and questions of the origins of resources, in terms of who is funding the work surrounding a piece of software (``the political'') and the manner in which resources come to the project initially (commercial, open source, closed partnerships, grant funded) and long-term (all four plus paid support).

@@

Another approach is to try to reduce the difficulty of citing software for authors, acknowledging that authors are often working with software that itself wraps other software and therefore hides that software. Knepley et al~\cite{Knepley_WSSSPE} approach this by proposing a mechanism by which the software itself, after it has run, provides the user with a set of citations that are relevant to the code actually run. They describe a prototype implementation whereby the citations are embedded in libraries and reported along with the results, via a commandline interface~\cite{Knepley_WSSSPE}. Discussion highlighted the difficulty that attempting to acknowledge the contributions of all pieces of dependent code within a paper faces the difficulty of creating very long citation lists, straining the analogy of code used to papers cited. Katz approaches this issue by proposing a system of ``transitive credit,'' recording dependencies and relative contributions outside particular papers, relieving authors from the responsibility of acknowledging each and every dependency. Instead authors would acknowledge the percentage contribution of the software they used directly and an external system would then be able to recursively allocate that credit to those who had provided dependencies. Finally Priem and Piwowar argued that machine learning techniques could examine the body of published literature and extract mentions of software, coping with the multitude of informal ways in which authors mention software they have used~\cite{Priem_WSSSPE}.  Discussion included turning the emphasis from asking users to improve their citation of software contributions, to ask how projects producing software might monitor the literature to improve their ability to show impact. Michael McLennan described the approach taken by the NanoHub project to scan the literature using keywords and names of known users to discover papers that are likely to have used their software and platform, then to have graduate students read each paper, highlighting any mention and perhaps following up with the authors to locate stories of impact.  Work since the workshop has described this process at \textit{publications.wikia.com}.

Acknowledgement in science does not come only in publications, of course. A key location for visibility might be in the grant funding process, both in bio-sketches and in grant project reporting. Some progress has been made here. Representatives of the NSF at the meeting emphasized these opportunities and encouraged participants to take advantage of them. Nonetheless, there was skepticism that peer review panels would value these contributions in the same ways as publications.

Priem and Piwowar argued for additionally looking beyond publications and drawing on evidence of contribution, and impact, recorded in other online systems, such as source code repositories like Github, both code contributions, downloads of code and dependencies as well as conversations about software (mailing lists, twitter and beyond)~\cite{Priem_WSSSPE}. They argue for providing scholars with flexible resources so that they can tell their own stories in the manner most appropriate for them and their audiences, a principle of the ``altmetrics'' approach.

@@

The workshop contributions in this group were concerned with the
aspect of how implementation of best practices and other
recommendations for improving scientific software sustainability could
be promoted. Specifically, if scientific software is to become more
sustainable, corresponding policies and guidelines need to be such
that the scientific community can follow and implement them. This is
considerably more challenging that it might seem at first, because in
the reality of science today resources, both financial and personnel,
that could be devoted to implementation are very limited, and the
reward system does not encourage scientists to do so. Furthermore,
implementing sustainability-targeting policies and guidelines often
takes a variety of specialized software engineering expertises, which
are not necessarily found in a single engineer, and much less so in a
domain scientist cross-trained in programming. Adding to the policy
implementation challenges, applicable sustainability-promoting
practices and guidelines will change through a software project's
lifecycle, in particular as it gains maturity.

Two of the papers in this group focus on specific facets of software
design that are important factors in a project's sustainability but
are often addressed only late in the scientific software development
cycle, if at all: Krintz et al~\cite{Krintz_WSSSPE} look at API
governance, and Heiland et al~\cite{Heiland_WSSSPE} discuss maturity
models for software security. The other two papers discuss
implementation strategies for science from the perspective of
facilitating many or all facets of sustainanability-oriented software
design for science: Blanton and Lenhardt\cite{Blanton_WSSSPE} contrast
large projects that have software infrastructue development built-in
with cross-training domain scientist PIs in software engineering best
practices. Huang and Lapp~\cite{Huang_WSSSPE} discuss how various
specialized software engineering skills could be turned into shared
instrumentation with low barriers to access.

Krintz et al~\cite{Krintz_WSSSPE} describe how in an era in which
computing frequently takes place in a distributed cloud, the control
over digital resources is increasingly shifting from physical
infrastructure to APIs, in particular web-service APIs. Yet, as Krintz
et al observe, although science communities have (for the most part)
developed solid practices for managing physical IT infrastructure in
data centers, corresponding practices and technology for API
governance, refered to by Krintz et al as the ``combined policy,
implementation, and deployment control'', is far less developed. As
Krintz et al demonstrate, web APIs can and do change, sometimes quite
frequently. raising the need to port dependent applications. The
effort required for porting is notoriously difficult to estimate,
making it nearly impossible for IT organizations to assess and thus
properly manage the impact of API changes. To address this, Krintz et
al propose a mechanism that evaluates the porting effort between two
versions of a web-service API in a formal and automated way. To
analyze the porting effort, they divide API changes into
\emph{syntactic similarity}, the changes in inputs and outputs, and
into \emph{semantic similarity}, the changes in the API's behavior. In
initial tests, their method showed good congruence with human
developers in scoring porting effort, offering the possibility that
API governance can become as solid a part of scientific IT management
as data center infrastructure management is today.

% Next, we’ll look at issues of software security.

% Heiland et al from the Center for Trustworthy Scientific Infrastructure discussed issues related to implementing strong security measures for scientific software. 
% They point out that cybersecurity is rarely addressed in scientific software design.

% Security considerations for software vary depending on the maturity level of the software package.
% But when we initially develop scientific software, we generally don’t know what the final maturity level will be.
% Scientific software developers are probably not aware of best practices for cybersecurity.
% So the authors introduce the concept of Software Security Maturity Models, such as OpenSAMM and BSI-MM. 
% These are used in industry to identify and define security vulnerabilities at different stages of the software life cycle.

% They suggest that a similar Software Security Maturity Model can formalize this process:
% It provides classification of software security practices.
% It provides a path for tightening security practices as a package’s maturity level increases.
% It emphasizes understandability over complexity.

% Finally, we’ll look at some papers that discuss implementing sustainability in scientific software.

% Blanton and Lenhardt from the Renaissance Computing Institute discuss these issues from a user perspective.

% The authors focus on a point that has been brought up many times in this context:
% There is a tension between writing code that is good enough just to “get it done,” i.e. to publish a paper about scientific results obtained using software, 
% and “getting it right,” that is, developing software that is comprehensible to future users and reviewers.
% Just because the elevator panel works like this doesn’t mean it’s sustainable for the long run.
% We don’t have a way to validate that the software used in a paper is actually done right.
% The best way to get software designed correctly is to make sure best practices are considered from the start.

% The authors highlight two models for sustainable software, at different extremes:
% One is what they call “co-funding”:
% In these projects, usually large, multi-year collaborations, 
% there is equal emphasis on both the science and the software development. 
% Both are planned into the project from inception. 
% In the life sciences, the iPlant Collaborative, Galaxy Project, and Qiime are good examples of these sorts of large, well-designed projects.

% At the other extreme, they discuss “software carpentry”: 
% in this model, it’s assumed that the scientists themselves will write and maintain their code.
% Groups like Software Carpentry and ROpenSci assume that 
% scientists won’t have access to dedicated software engineering, 
% so they try to give them tools to use best practices in their own software development.

% There might be a middle ground here: a way to get the engineering expertise that large co-funded projects have to individual scientist-developers.
% Hilmar Lapp of NESCent and I discuss one such possibility in our paper, Software Engineering as Instrumentation for the Long Tail of Scientific Software.

% What do we mean when we refer to the “long tail” of scientific software?
% Think of the distribution of resources in scientific software. Most are focused on big projects with lots of community buy-in and funding. But a lot of scientific software exists away from this model.
% For example, scientific software can be used long after the original developer has moved on or the funding runs out.
% Look at MacClade: it was originally released in 1986 and last updated in 2005,
% but it was still cited over 400 times in 2013!
% The scientists who developed it have a newer package, Mesquite, that was meant to replace MacClade, but they haven’t had sufficient time or resources to maintain either package fully, let alone both of them.

% Another dimension of the long tail can also be found in my particular research domain. 
% In the field of phylogenetics, we have a lot of programs that implement different computational methods in slightly different ways.
% Here, Joe Felsenstein has listed some (but not anywhere near all) phylogenetics packages available online.
% Most of these programs are developed by academic scientists…
% They generally have limited training in software engineering
% Limited time or career incentive to improve software
% Limited funding

% So, to summarize a bit:
% Making sustainable software means we have to pay attention to many facets of software design, like APIs, security, user experience, testing, etc.
% A single project that requires one full­-time software engineer may actually require fractions of different kinds of engineers. 
% But long-tail projects can’t even fund one FTE, let alone one that can address all these facets.

% Then we have to consider that the users of scientific software are scientists, 
% so the developers need to understand the users and the science.
% This is the idea of a “t-skilled” person: 
% one who is both well-versed in a scientific domain 
% and deeply experienced in one or more facets of software engineering. 
% These people are pretty rare in the first place and difficult to retain in academia, because the academic career structure doesn’t incentivize this.

% We should look at software engineering as an expensive resource, but one that needs to be accessible to scientists at all levels.
% Think of it as analogous to DNA sequencing:
% Sequencers used to be something that individual labs and institutions had to 
% buy, maintain, and operate themselves, 
% so only highly-funded operations had them and probably didn’t use them to their full capacity even when they had one.
% But now, core facilities provide the instrumentation and service to labs of any size. 
% Anyone can pay a core facility to sequence their samples for them and provide quality control and bioinformatics advice as additional services.

% We propose that software engineering can be “instrumented” in a similar way. 
% Let’s create a nonprofit center for scientific software engineering.
% This center can hire these t-skilled personnel and provide access to them for projects at contracted cost.
% Because the center is focused on providing development services to scientific projects, it is not tied to the long-term success or failure of any individual project.
% It would emphasize the centrality of doing good science by making functional software tools as envisioned by scientists.

% So, to conclude…
% Implementing policies to encourage sustainability in scientific software 
% requires that many facets of good software design are addressed throughout the lifecycle of these projects. 
% But most of them aren’t addressed in the status quo.
% We’ve highlighted some of these facets today and suggested some possible solutions.
% Large projects can afford to hire software engineers with the expertise to implement these facets correctly.
% Grassroots developer groups can provide guidance to scientists about best practices in software development.
% We think there is a place for a software engineering center that can provide 
% both engineering expertise and guidance 
% with a contract-driven instrumentation model 
% to the scientific software in the long tail.

@@

\item Chandra Krintz, Hiranya Jayathilaka, Stratos
  Dimopoulos, Alexander Pucher, Rich Wolski. Developing Systems for API Governance~\cite{Krintz_WSSSPE}

@@

% Below: What are the goals of this discussion? TODO: Better motivate. May be make a brief list of items discussed in the workshop.
We aim to understand the practicalities of the key points discussed in the
workshop, and hope to extract the best practices associated with the case
studies. We discuss how the issues are applicable to the cases and try to
understand how the course of current and future development projects might be
altered to implement lessons learned. In particular, we discuss the following issues:
\begin{itemize}
    \item How closely are the best practices discussed in the workshop were followed?
    \item Career path: Are the original developers still around?
    \item What conscious sustainability efforts were undertaken by stakeholders?
    \item What is the dominant sustainability factor: science, publication, (re)usability?
\end{itemize}

@@

Some examples of such software discussed as part of the workshop are,
\emph{DUNE}, \emph{R/qtl}, \emph{Kitware}, \emph{PETSc}, \emph{MINRES-QLP},
\emph{FASTMath}.  Most of the above efforts are focused on one scientific
and/or applied mathematics domain. However, sometimes such projects grow beyond
the initial vision of developers. One such example is \emph{Kitware}, which
while being a software product specializing in the scientific process, has a
core focus of developing communities around software processes. An instance of
this process is the development of the \emph{CMake} build utility, which
started out as a building tool for \emph{ITK} but grew to become a generic
build utility for C++ projects. Similarly, \emph{PETSc} is growing towards
becoming a general purpose utility system usable for solving variety of
scientific problems.
%

%Summary of above discussed cases: lessons, trends, commonalities, is there an ideal sustainability scenario?
In order to investigate the much discussed issue of career paths for developers
of scientific software, we conducted a small survey of the use-cases discussed
above with respect to the current status of its core developers. We looked at
the workshop submissions and the original cited papers for the respective
software. Where available, we looked at the webpages and source code of the
earliest versions to find the original authors. We excluded relatively new
projects such as \emph{SciPy} and \emph{numFOCUS} from this survey. we find
that in a majority of scientific software more than one original developers are
still around, in some cases for more than 10 years, (eg. \emph{R/qtl}).
Whereas, in the case of utility software projects, it appears that at least one
original developer is still around often in a part-time role. In other cases,
it appears that the original developers have went on to assume a larger role,
again with the same project.

From this survey, it appears that the career paths for utility software
developers are broader compared to the developers of niche software. However,
as projects as a whole, the small-scale, specific software have a lower risk
and cost in terms of sustaining as compared to utility software. Often the
sustainability of such projects is achieved by the fact that the core
developer or team heavily utilizes the software for their science, eg.
\emph{R/qtl, PETSc}. Furthermore, the development of scientific software requires more
scientific skills compared with those of utility software. This diminishes the
need of specialist programmers and practices in many cases where the bulk of
development is carried out by a domain scientist.

>>>>

One group of papers presented frameworks that were primarily about characteristics of software artifacts, connecting with the long discourse on software quality. This approach is realized in adjectives that can be applied to pieces of software but might also extend to describe software projects.  Thus Calero and colleagues propose adding elements to the ISO standards for measuring software quality: proposing an additional dimension of quality they call perduability or ability to ``last over time'' made up of three characteristics: reliability, maintainability and adaptability. This shows some overlap with the framework employed by Venters et al.~\cite{Venters_WSSSPE} who employ the adjectives of ``extensibility, interoperability, maintainability, portability, reusability and scalability,'' anticipating the sorts of work that people would need to do with a software artifact in the future, ``as stakeholders requirements, technology and environments evolve and change.'' Venters et al. argue that these choices are really made very early because they relate to the architecture of the software and involve trade offs that ought to be analyzed alongside other attributes of the software artifact. Lenhardt et al compare the software lifecycle to the data lifecycle to argue for the inclusion of metadata throughout a piece of software's life (discussing, for example, how it has been built and tested and what ``data in'' and ``data out'' has been considered.) In addition their analogy suggests that the software lifecycle might add a phase of ``preservation'' and draw on understandings of what that involves from studies of data.  In sum, then, these frameworks focused on what needs to be accomplished to have more sustainable software.

A second theme in these papers was the continued availability of resources to accomplish the goals of sustainability. The elements of these frameworks focused far more on the organization of a software project than they did on characteristics of the artifact itself (although it is certainly true that the adjectives discussed above could be applied to a software project). For example Pierce et al focus on the way that the project is run, particularly in terms of how those involved communicate and jointly set priorities, a process they call governance. In particular they argue that because sustainability relates to having ongoing resources, governance must be open to receive diverse input (by occurring online, asynchronously, and publicly) and thus have the potential to ``transform passive users into active contributors''. They argue that the Apache Software Foundation's incubation process teaches this and could be learned from by projects throughout scientific software.  Katz and Proctor also discuss governance, describing two modes: ``top-down'' and ``bottom-up'' governance~\cite{Katz_WSSSPE}.  They place governance alongside questions of the artifact (``technical''), and questions of the origins of resources, in terms of who is funding the work surrounding a piece of software (``the political'') and the manner in which resources come to the project initially (commercial, open source, closed partnerships, grant funded) and long-term (all four plus paid support).

@@

Another approach is to try to reduce the difficulty of citing software for authors, acknowledging that authors are often working with software that itself wraps other software and therefore hides that software. Knepley et. al. approach this by proposing a mechanism by which the software itself, after it has run, provides the user with a set of citations that are relevant to the code actually run. They describe a prototype implementation whereby the citations are embedded in libraries and reported along with the results, via a commandline interface~\cite{Knepley_WSSSPE}. Discussion highlighted the difficulty that attempting to acknowledge the contributions of all pieces of dependent code within a paper faces the difficulty of creating very long citation lists, straining the analogy of code used to papers cited. Katz approaches this issue by proposing a system of ``transitive credit,'' recording dependencies and relative contributions outside particular papers, relieving authors from the responsibility of acknowledging each and every dependency. Instead authors would acknowledge the percentage contribution of the software they used directly and an external system would then be able to recursively allocate that credit to those who had provided dependencies. Finally Priem and Piwowar argued that machine learning techniques could examine the body of published literature and extract mentions of software, coping with the multitude of informal ways in which authors mention software they have used~\cite{Priem_WSSSPE}.  Discussion included turning the emphasis from asking users to improve their citation of software contributions, to ask how projects producing software might monitor the literature to improve their ability to show impact. Michael McLennan described the approach taken by the NanoHub project to scan the literature using keywords and names of known users to discover papers that are likely to have used their software and platform, then to have graduate students read each paper, highlighting any mention and perhaps following up with the authors to locate stories of impact.  Work since the workshop has described this process at \textit{publications.wikia.com}.

Acknowledgement in science does not come only in publications, of course. A key location for visibility might be in the grant funding process, both in bio-sketches and in grant project reporting. Some progress has been made here. Representatives of the NSF at the meeting emphasized these opportunities and encouraged participants to take advantage of them. Nonetheless, there was skepticism that peer review panels would value these contributions in the same ways as publications).

Priem and Piwowar argued for additionally looking beyond publications and drawing on evidence of contribution, and impact, recorded in other online systems, such as source code repositories like github, both code contributions, downloads of code and dependencies as well as conversations about software (mailing lists, twitter and beyond)~\cite{Priem_WSSSPE}. They argue for providing scholars with flexible resources so that they can tell their own stories in the manner most appropriate for them and their audiences, a principle of the ``alt.metrics'' approach.

@@

@@

\item Rich Wolski, Chandra Krintz, Hiranya Jayathilaka, Stratos
  Dimopoulos, Alexander Pucher. Developing Systems for API Governance~\cite{Wolski_WSSSPE}

@@

% Below: What are the goals of this discussion? TODO: Better motivate. May be make a brief list of items discussed in the workshop.
We aim to understand the practicalities of the key points discussed in the
workshop, and hope to extract the best practices associated with the case
studies. We discuss how the issues are applicable to the cases and try to
understand how the course of current and future development projects might be
altered to implement lessons learned. We discuss vatious issues including
career paths of original developers and investigate if the original developers
are still around?

@@

Some examples of such software discussed as part of the workshop are,
\emph{DUNE}, \emph{R/qtl}, \emph{Kitware}, \emph{PETSc}, \emph{MINRES-QLP},
\emph{FASTMath}.  Most of the above efforts are focused on one scientific
and/or applied mathematics domain. However, sometimes such projects grow beyond
the initial vision of developers. One such example is \emph{Kitware}, which
while being a software product specializing in the scientific process, has a
core focus of developing communities around software processes. An instance of
this process is the development of the \emph{CMake} build utility, which
started out as a building tool for \emph{ITK} but grew to become a generic
build utility for C++ projects. Similarly, \emph{PETSc} is growing towards
becoming a general purpose utility system usable for solving variety of
scientific problems.
% 
In order to investigate the much discussed issue of career paths for developers
of scientific software, we conducted a small survey of the use-cases discussed
above with respect to the current status of its core developers. We looked at
the workshop submissions and the original cited papers for the respective
software. Where available, we looked at the webpages and source code of the
earliest versions to find the original authors. We excluded relatively new
projects such as \emph{SciPy} and \emph{numFOCUS} from this survey. we find
that in a majority of scientific software more than one original developers are
still around, in some cases for more than 10 years, (eg. \emph{R/qtl}).
Whereas, in the case of utility software projects, it appears that at least one
original developer is still around often in a part-time role. In other cases,
it appears that the original developers have went on to assume a larger role,
again with the same project.

In conclusion, there are clear distinctions in the characteristics and the
kinds of challenges faced between generic utility oriented and specific
scientific domain oriented software projects. In terms of career paths of
original developers, it appears that the career paths for utility software
developers are broader compared to the developers of scientific software.
However, it appears that in the case of scientific software, core competencies
of original developers tend to be domain science who happen to be the coders of
their software. As projects as a whole, the small-scale, specific software have
a lower risk and cost in terms of sustaining as compared to utility software.
Often the sustainability of such projects is achieved by the fact that the core
developer or team heavily utilizes the software for their science, eg.
\emph{R/qtl, PETSc}. Furthermore, the development of scientific software
requires more scientific skills compared with those of utility software. This
diminishes the need of specialist programmers and practices in many cases where
the bulk of development is carried out by a domain scientist. With these
inferences, we believe that separate guidelines and/or principles for
sustainability be defined for each category of software projects in science and
they be given separate treatment.
